---
title: "Mini Data Analysis Milestone 2"
output: github_document
---

*To complete this milestone, you can either edit [this `.rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are commented out with `<!--- start your work here--->`. When you are done, make sure to knit to an `.md` file by changing the output in the YAML header to `github_document`, before submitting a tagged release on canvas.*

# Welcome to the rest of your mini data analysis project!

In Milestone 1, you explored your data. and came up with research questions. This time, we will finish up our mini data analysis and obtain results for your data by: 

- Making summary tables and graphs 
- Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

We will also explore more in depth the concept of *tidy data.* 

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class.

# Instructions

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work here--->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to your mini-analysis GitHub repository, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 50 points: 45 for your analysis, and 5 for overall reproducibility, cleanliness, and coherence of the Github submission. 

**Research Questions**: In Milestone 1, you chose two research questions to focus on. Wherever realistic, your work in this milestone should relate to these research questions whenever we ask for justification behind your work. In the case that some tasks in this milestone don't align well with one of your research questions, feel free to discuss your results in the context of a different research question.

# Learning Objectives

By the end of this milestone, you should:

-   Understand what *tidy* data is, and how to create it using `tidyr`.
-   Generate a reproducible and clear report using R Markdown.
-   Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

# Setup

Begin by loading your data and the tidyverse package below:

```{r, message = FALSE}
library(datateachr) # <- might contain the data you picked!
library(tidyverse)
library(broom)
library(gapminder)
library(tsibble)
library(testthat)
library(digest)
library(ggridges)
library(scales)
library(rlang)
library(readr)
library(tidymodels)
```

# Task 1: Process and summarize your data 

From milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions. 

### 1.1 (1 point) 

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:

<!-------------------------- Start your work below ---------------------------->

1. How does the *height_range_id* relate to the *diameter*? For example, are the larger diameter trees in a taller height group? 
  
2. What types of species are in the neighbourhoods where UBC campus is located? 
  
3. How does *street_side_name* relate to *curb*? For example, does the odd numbered side of the street have more trees next to a curb than the even numbered side? 

4. Which neighbourhood on average has the largest trees by diameter? 

<!----------------------------------------------------------------------------->

Here, we will investigate your data using various data manipulation and graphing functions.

### 1.2 (8 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question.

Ensure that the output of each operation is printed!

Also make sure that you're using dplyr and ggplot2 rather than base R. Outside of this project, you may find that you prefer using base R functions for certain tasks, and that's just fine! But part of this project is for you to practice the tools we learned in class, which is dplyr and ggplot2.

**Summarizing:**

1.  Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2.  Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3.  Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4. Compute the proportion and counts in each category of one categorical variable across the groups of another categorical variable from your data. Do not use the function `table()`!

**Graphing:** 

6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
7. Make a graph where it makes sense to customize the alpha transparency.

Using variables and/or tables you made in one of the "Summarizing" tasks: 

8. Create a graph that has at least two geom layers. 
9. Create 3 histograms, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->

## Question 1: How does the *height_range_id* relate to the *diameter*? For example, are the larger diameter trees in a taller height group?

### Summarizing 

Here I will compute the number of observations in *height_range_id*, a categorical variable. This does not fully answer the question, but it helps to understand the the *height_range_id* variable. 

```{r}

Q1_summary <- count(vancouver_trees, height_range_id) #this is making a table counting the number of observations in each group for the height_range_id variable
print(Q1_summary) #this is printing the tibble  

```

### Graphing 

Here I will make a jitterplot graph comparing *diameter* across *height_range_id*, where it makes sense to customize the alpha transparency due to the density of the points. This is essentially answering the question, and shows me which height range has the largest diameter trees. 

```{r}

Q1_graph <- ggplot(vancouver_trees, aes(height_range_id, diameter)) + #this is specifying the variables for graphing
    geom_jitter(alpha = 0.1, width = 0.1) #this is specifing the type of graph, jitterplot, the transparency of the points, and the width of the sections for each height range  

print(Q1_graph) #this is printing the graph 

```

## Question 2: What types of species are in the neighbourhoods where UBC campus is located? 

### Summarizing 

Here I will compute the proportion and counts in each category of *species_name*, a categorical variable, across the groups of *neighbourhood_name*, another categorical variable. This does not fully answer the question about the neighbourhood where UBC campus is located, but it is starting to look at the relationship between species and neighbourhoods. 

```{r}

Q2_summary <- vancouver_trees %>%    
    group_by(neighbourhood_name, species_name) %>%  #this is specifying the variables, and their order of consideration 
    summarise (n = n()) %>% #this is creating the new column n, which counts the number of observations of each species in each neighbourhood 
    mutate(freq = n / sum(n)) #this is creating the new column freq, which is the proportion of each species in each neighbourhood 

print(Q2_summary) #this is printing the tibble 

```

### Graphing 

Here I will graph the number of trees in each category of *species_name*. I will make a bar graph with a log scale on the y axis, so that the lower values will not appear as zero values. I will relabel the axes to make the variables more clear to a viewer. This will not fully answer the question, but it starts to give an idea of just how many species are present in vancouver and the wide range of abundances they are present in.  

```{r}

Q2_graph <- ggplot(vancouver_trees, aes(species_name)) + #this is specifying the variable on the x axis 
  geom_bar() + #this is specifying that ut will be a bar graph 
  scale_y_log10() +
  labs(x = "Tree Species", y = "Number of Trees") #this is renaming the axis labels 
    
print(Q2_graph) #this is printing the graph 

```

## Question 3: How does *street_side_name* relate to *curb*? For example, does the odd numbered side of the street have more trees next to a curb than the even numbered side?

### Summarizing 

Here I will compute the proportion and counts in each category of *curb*, a categorical variable, across the groups of *street_side_name*, another categorical variable. This essentially answers the research question, however it would be better visualized in a graphical form. 

```{r}

Q3_summary <- vancouver_trees %>%    
    group_by(street_side_name, curb) %>%  #this is specifying the variables, and their order of consideration 
    summarise (n = n()) %>% #this is creating the new column n, which counts the number of observations of each curb type (yes or no) on each street side type 
    mutate(freq = n / sum(n)) #this is creating the new column freq, which is the proportion of each curb type and street side type 

print(Q3_summary) #this is printing the tibble 

```

### Graphing 

Here I will graph the number of trees in each category of *street_side_name* and *curb*. I will make a bar graph with a log scale on the y axis, so that the lower values will not appear as zero values. I will relabel the axes and the legend tile to make the variables more clear to a viewer. This will essentially answer the question, and I will be able to see whether the presence or absence of a curb has more or less trees in each street side type. 

```{r}

Q3_graph <- ggplot(vancouver_trees, aes(street_side_name)) + #this is specifying the x axis variable 
  geom_bar(aes(fill = curb), position = "dodge") + #this is specifying that the x axis variable will be split into the categories for the curb type (yes or no), and this is creating the bar graph 
  scale_y_log10() + #this is creating a log scale on the y axis 
  labs(x = "Side of the Street", y = "Number of Trees", fill = "Curb Presence") #this is renaming the axis and legend labels 
    
print(Q3_graph) #this is printing the graph 

```

## Question 4: Which neighbourhood on average has the largest trees by diameter?

### Summarizing

Here I will compute the number of observations in *neighbourhood_name*, a categorical variable. This does not fully answer the question, but it helps to understand the *neighbourhood_name* variable. 

```{r}

Q4_summary <- count(vancouver_trees, neighbourhood_name) #this is making a table counting the number of observations in each group for the neighbourhood_name variable

print(Q4_summary) #this is printing the tibble  

```

### Graphing 

Here I will make a jitterplot graph comparing *diameter* across *neighbourhood_name*, where it makes sense to customize the alpha transparency due to the density of the points. This is essentially answering the question, and shows me which neighbourhood has the largest diameter trees. 

```{r}

Q4_graph <- ggplot(vancouver_trees, aes(neighbourhood_name, diameter)) + #this is specifying the variables for graphing
    geom_jitter(alpha = 0.1, width = 0.1) #this is specifing the type of graph, jitterplot, the transparency of the points, and the width of the sections for each neighbourhood name 

print(Q4_graph) #this is printing the graph 

```

'''{r}
<!----------------------------------------------------------------------------->

### 1.3 (2 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!------------------------- Write your answer here ---------------------------->
## Progress on Research Questions 

1. How does the *height_range_id* relate to the *diameter*? For example, are the larger diameter trees in a taller height group? 
  * This question has essentially been answered with the jitterplot graph. It had interesting results, as the largest diameter tree actually only falls in the 3rd heught group (out of 10 groups). 
  
2. What types of species are in the neighbourhoods where UBC campus is located? 
  * This question has not yet been answered, and I am excited to explore it more! 
  * I have realised that UBC specifically is not included in this dataset, but I have decided to narrow it to one of the closest neighbourhoods to UBC, which is West Point Grey
  *Therefore, I would like to refine this question into a new question: 
    * What types of species are in the West Point Grey neighbourhood?
  
3. How does *street_side_name* relate to *curb*? For example, does the odd numbered side of the street have more trees next to a curb than the even numbered side?
  * This question has essentially been answered with the bar graph, and it appears that the odd and equal side of the street have very similar proportions of curb and non-curb trees. And overall there are more trees that are on a curb than are not, across all street side types. 

4. Which neighbourhood on average has the largest trees by diameter? 
  * This question has essentially been answered, and the largest tree is visible in one neighbourhood on the jitterplot graph.
  * I would like to refine this questions further into a new question: 
    * What is the largest tree species by diameter in the West Point Grey neighbourhood? 
 

<!----------------------------------------------------------------------------->

# Task 2: Tidy your data 

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

-   Each row is an **observation**
-   Each column is a **variable**
-   Each cell is a **value**

### 2.1 (2 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have \>8 variables, just pick 8, and explain whether the data is untidy or tidy.

<!--------------------------- Start your work below --------------------------->

## Tidy Data 

Since *vancouver_trees* has more than 8 variables, I will be looking at the first 8 variables for this question. These are *tree_id*, *civic_number*, *std_street*, *genus_name*, *species_name*, *cultivar_name*, *common_name*, and *assigned*. Based on these 8 variables, the data is tidy. Each row is a single observation, each column is a single variable, and each cell is a single value. 

```{r}

head(vancouver_trees) #this is to take a look at the columns and variables 

```

<!----------------------------------------------------------------------------->

### 2.2 (4 points)

Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

<!--------------------------- Start your work below --------------------------->

## Untidying the Data 

Here I will untidy the data by combining the columns *species_name* and *common_name* into one column. I will then retidy the data and separate them back into their original columns 

```{r}

print(head(vancouver_trees)) #this is to show the data in it's original state 

```
```{r}

 untidy_data <- vancouver_trees %>%
    unite(col = species_and_common_name, c("species_name", "common_name"), sep = ", ") #this is combining the two columns, renaming the new column, and making the values separated by a comma

print(untidy_data) #this is printing the tibble of untidy data 

```

```{r}

 tidy_data <- untidy_data %>%
    separate(col = species_and_common_name, c("species_name", "common_name"), sep = ", ") #this is combining the two columns, renaming the new column, and making the values separated by a comma

print(tidy_data) #this is printing the tibble of tidy data 

```


<!----------------------------------------------------------------------------->

### 2.3 (4 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in the remaining tasks:

<!-------------------------- Start your work below ---------------------------->

1.  What types of species are in the West Point Grey neighbourhood?
2.  What is the largest tree species by diameter in the West Point Grey neighbourhood? 

<!----------------------------------------------------------------------------->

Explain your decision for choosing the above two research questions.

<!--------------------------- Start your work below --------------------------->

I chose the above two research questions because they both remain unanswered, and because I am genuinely interested in the trees around UBC campus! 

<!----------------------------------------------------------------------------->

Now, try to choose a version of your data that you think will be appropriate to answer these 2 questions. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidy'ing, dropping irrelevant columns, etc.).

(If it makes more sense, then you can make/pick two versions of your data, one for each research question.) 

<!--------------------------- Start your work below --------------------------->

## New Tibble from *vancouver_trees*

I will be making a version of the *vancouver_trees* dataset which better suits the analysis for my research questions. I will be utilizing the select(), filter(), arrange(), and group_by() functions to achieve this. 

```{r}

edited_vancouver_trees <- vancouver_trees %>% #this is creating the new tibble
  select(tree_id, common_name, neighbourhood_name, diameter) %>% #this is removing all columns except for these 4 I am interested in 
  filter(neighbourhood_name == "WEST POINT GREY") %>% #this is filtering the neighbourhood_name to only include the West Point Grey neighbourhood
  arrange(min_rank(desc(diameter))) %>% #this is arranging the data in descending order of the diameter, so the largest diameter is first 
  group_by(common_name) #this is grouping by common name, so that trees with the same diameter will then be ordered alphabetically by common name
  
print(edited_vancouver_trees) #this is printing the new tibble 

```

<!----------------------------------------------------------------------------->

# Task 3: Modelling

## 3.0 (no points)

Pick a research question from 1.2, and pick a variable of interest (we'll call it "Y") that's relevant to the research question. Indicate these.

<!-------------------------- Start your work below ---------------------------->

**Research Question**: What is the largest tree species by diameter in the West Point Grey neighbourhood?

**Variable of interest**: diameter  

<!----------------------------------------------------------------------------->

## 3.1 (3 points)

Fit a model or run a hypothesis test that provides insight on this variable with respect to the research question. Store the model object as a variable, and print its output to screen. We'll omit having to justify your choice, because we don't expect you to know about model specifics in STAT 545.

-   **Note**: It's OK if you don't know how these models/tests work. Here are some examples of things you can do here, but the sky's the limit.

    -   You could fit a model that makes predictions on Y using another variable, by using the `lm()` function.
    -   You could test whether the mean of Y equals 0 using `t.test()`, or maybe the mean across two groups are different using `t.test()`, or maybe the mean across multiple groups are different using `anova()` (you may have to pivot your data for the latter two).
    -   You could use `lm()` to test for significance of regression coefficients.

<!-------------------------- Start your work below ---------------------------->

## Fitting a Model 

Here I will fit a model that makes predictions on *diameter* based on *common_name*, using the lm() function. 

```{r}

model_fit <- lm(diameter ~ common_name, edited_vancouver_trees) #this is fitting a model that makes predictions on diameter based on common name

print(model_fit) #this is printing the output of the model 

```


<!----------------------------------------------------------------------------->

## 3.2 (3 points)

Produce something relevant from your fitted model: either predictions on Y, or a single value like a regression coefficient or a p-value.

-   Be sure to indicate in writing what you chose to produce.
-   Your code should either output a tibble (in which case you should indicate the column that contains the thing you're looking for), or the thing you're looking for itself.
-   Obtain your results using the `broom` package if possible. If your model is not compatible with the broom function you're needing, then you can obtain your results by some other means, but first indicate which broom function is not compatible.

<!-------------------------- Start your work below ---------------------------->

## Predictions on Diameter 

From the *edited_vancouver_trees* tibble, I can see that the Western Red Cedar is the largest species by diameter in the West Point Grey neighbourhood. I am curious what the overall expected diameter is of the Western Red Cedar in this neighbourhood, and I will use my model to predict this. 


```{r}

augment(model_fit, newdata = tibble(common_name = "WESTERN RED CEDAR")) %>% #This is testing the model specifically on the Western Red Cedar
  print(n = 1) #this is printing the tibble 

```

<!----------------------------------------------------------------------------->

# Task 4: Reading and writing data

Get set up for this exercise by making a folder called `output` in the top level of your project folder / repository. You'll be saving things there.

## 4.1 (3 points)

Take a summary table that you made from Task 1, and write it as a csv file in your `output` folder. Use the `here::here()` function.

-   **Robustness criteria**: You should be able to move your Mini Project repository / project folder to some other location on your computer, or move this very Rmd file to another location within your project repository / folder, and your code should still work.
-   **Reproducibility criteria**: You should be able to delete the csv file, and remake it simply by knitting this Rmd file.

<!-------------------------- Start your work below ---------------------------->

## Writing a csv File 

Here I will take the summary table *Q1_smmary* from Task 1, and write it as a cvs file in my "output" folder. 

```{r}

write_csv(Q1_summary, here::here("Output", "Q1_summary.csv")) #writes a cvs file of the Q1_summary table into the Output folder 

```

<!----------------------------------------------------------------------------->

## 4.2 (3 points)

Write your model object from Task 3 to an R binary file (an RDS), and load it again. Be sure to save the binary file in your `output` folder. Use the functions `saveRDS()` and `readRDS()`.

-   The same robustness and reproducibility criteria as in 4.1 apply here.

<!-------------------------- Start your work below ---------------------------->

## Writing an R Binary File

Here I will take the *model_fit* and write it to an R binary file, load it again and save it to the output folder. 

```{r}

saveRDS(model_fit, here::here("Output", "model_fit.rds")) #write the model to an R binary file and save into the Output folder 

readRDS(here::here("Output", "model_fit.rds"))

```
<!----------------------------------------------------------------------------->

# Overall Reproducibility/Cleanliness/Coherence Checklist 

Here are the criteria we're looking for.

## Coherence (0.5 points)

The document should read sensibly from top to bottom, with no major continuity errors. 

The README file should still satisfy the criteria from the last milestone, i.e. it has been updated to match the changes to the repository made in this milestone. 

## File and folder structure (1 points)

You should have at least three folders in the top level of your repository: one for each milestone, and one output folder. If there are any other folders, these are explained in the main README.

Each milestone document is contained in its respective folder, and nowhere else.

Every level-1 folder (that is, the ones stored in the top level, like "Milestone1" and "output") has a `README` file, explaining in a sentence or two what is in the folder, in plain language (it's enough to say something like "This folder contains the source for Milestone 1").

## Output (1 point)

All output is recent and relevant:

-   All Rmd files have been `knit`ted to their output md files. 
-   All knitted md files are viewable without errors on Github. Examples of errors: Missing plots, "Sorry about that, but we can't show files that are this big right now" messages, error messages from broken R code
-   All of these output files are up-to-date -- that is, they haven't fallen behind after the source (Rmd) files have been updated.
-   There should be no relic output files. For example, if you were knitting an Rmd to html, but then changed the output to be only a markdown file, then the html file is a relic and should be deleted.

Our recommendation: delete all output files, and re-knit each milestone's Rmd file, so that everything is up to date and relevant.

## Tagged release (0.5 point)

You've tagged a release for Milestone 2. 

### Attribution

Thanks to Victor Yuan for mostly putting this together.
